---
title: "LiveImagingAnalysis_test"
output: html_document
date: "2026-03-17"
---

```{r setup, include=FALSE}
library(gridExtra)
library(reshape2)
library(LSD)
library(MASS)
library(ggplot2)
library(corrplot)
library( dplyr )
library( plotly )
library( DT )    
library(magrittr)
library( viridis )
library( pheatmap )
library( ggrepel )
library( tidyr )
library(tidyverse)
library( rbioapi )
library( ggthemes )
library(zoo)
library(corrplot)
library(glmnet)


source( "/Users/talfis/Desktop/Tal/R_scripts/AUXI.R" )
```

```{r}
    draw_it <- function( df, P1="", P2="", title = "" ) {
  dftmp = df[is.finite(df[[P1]]) & is.finite(df[[P2]]),]
  dftmp$dens1 = get_density(dftmp[[P1]],dftmp[[P2]], n = 100)
  ggplot(dftmp, aes(dftmp[[P1]],dftmp[[P2]])) +
  geom_point(aes(color = dens1)) +
  scale_color_viridis() +
  labs(x = P1, y = P2) + 
  ggtitle(paste0(title, " Pearson R = ", signif(cor(dftmp[[P1]],dftmp[[P2]], use = "complete.obs"),3), ", Spearman R = ", signif(cor(dftmp[[P1]],dftmp[[P2]], use = "complete.obs", method = "spearman"),3), ",\n number of genes = ",  ... = nrow(dftmp))) + theme(legend.position = "none")
    }

roll_mean_partial <- function(x, k) {
  rollapply(
    x,
    width = k,
    FUN = function(v) {
      if (all(is.na(v))) return(NA_real_)
      mean(v, na.rm = TRUE)
    },
    fill = NA,
    align = "center",
    partial = TRUE
  )
}

plot_metric = function(data, y_col, y_label, title_text) {
  ggplot(data, aes(x = `T..sec.`, y = .data[[y_col]], color = factor(`Track.ID`), group = `Track.ID`)) +
    geom_line(linewidth = 0.7, alpha = 0.9) +
    geom_point(size = 0.8, alpha = 0.7) +
    labs(
      title = title_text,
      x = "Time (sec)",
      y = y_label,
      color = "Track ID"
    ) + xlim(0, NA) +
    theme_bw() +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "none"
    )
}

ellipse_score <- function(x_pt, y_pt, x0, y0, long_axis, short_axis, angle_rad) {
  a <- long_axis / 2
  b <- short_axis / 2
  
  if (is.na(a) || is.na(b) || is.na(x0) || is.na(y0) || is.na(angle_rad) || a <= 0 || b <= 0) {
    return(NA_real_)
  }
  
  dx <- x_pt - x0
  dy <- y_pt - y0
  
  x_rot <-  dx * cos(angle_rad) + dy * sin(angle_rad)
  y_rot <- -dx * sin(angle_rad) + dy * cos(angle_rad)
  
  (x_rot^2)/(a^2) + (y_rot^2)/(b^2)
}

plot_mean_sd <- function(data, mean_col, sd_col, ylab, title) {

  ggplot(data, aes(x = T..sec., y = .data[[mean_col]])) +
    geom_ribbon(
      aes(
        ymin = .data[[mean_col]] - .data[[sd_col]],
        ymax = .data[[mean_col]] + .data[[sd_col]]
      ),
      alpha = 0.25
    ) +
    geom_line(linewidth = 1) +
    theme_bw() +
    labs(
      x = "Time (sec)",
      y = ylab,
      title = title
    )
}

zscore_safe <- function(v) {
  s <- sd(v, na.rm = TRUE)
  m <- mean(v, na.rm = TRUE)
  if (is.na(s) || s == 0) return(rep(NA_real_, length(v)))
  (v - m) / s
}

mean_sem <- function(x) {
  n <- sum(!is.na(x))
  if (n == 0) return(c(mean = NA_real_, sem = NA_real_))
  c(
    mean = mean(x, na.rm = TRUE),
    sem = sd(x, na.rm = TRUE) / sqrt(n)
  )
}

plot_avg_ccf <- function(data,
                         feature_x,
                         feature_y,
                         track_col = "Track.ID",
                         time_col = "cell_Frame",
                         lag_max = 20,
                         min_points = 10,
                         zscore = TRUE) {
  
  ccf_per_track <- data %>%
    arrange(.data[[track_col]], .data[[time_col]]) %>%
    group_by(.data[[track_col]]) %>%
    group_split() %>%
    map_dfr(function(df) {
      
      df <- df %>% arrange(.data[[time_col]])
      
      x <- df[[feature_x]]
      y <- df[[feature_y]]
      
      if (zscore) {
        x <- zscore_safe(x)
        y <- zscore_safe(y)
      }
      
      keep <- complete.cases(x, y)
      
      if (sum(keep) < min_points) {
        return(NULL)
      }
      
      cc <- ccf(x[keep], y[keep], lag.max = lag_max, plot = FALSE)
      
      tibble(
        track_id = unique(df[[track_col]])[1],
        lag = as.numeric(cc$lag),
        ccf = as.numeric(cc$acf)
      )
    })
  
  ccf_summary <- ccf_per_track %>%
    group_by(lag) %>%
    summarise(
      mean_ccf = mean(ccf, na.rm = TRUE),
      sem_ccf = sd(ccf, na.rm = TRUE) / sqrt(sum(!is.na(ccf))),
      n_tracks = sum(!is.na(ccf)),
      .groups = "drop"
    )
  
  p <- ggplot(ccf_summary, aes(x = lag, y = mean_ccf)) +
    geom_line() +
    geom_ribbon(
      aes(ymin = mean_ccf - sem_ccf,
          ymax = mean_ccf + sem_ccf),
      alpha = 0.3
    ) +
    theme_bw() +
    labs(
      title = paste("Average cross-correlation:", feature_x, "vs", feature_y),
      x = "Lag",
      y = "Cross-correlation"
    )
  
  list(
    per_track = ccf_per_track,
    summary = ccf_summary,
    plot = p
  )
}


safe_first <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  x[1]
}

safe_last <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) == 0) return(NA_real_)
  x[length(x)]
}

safe_sd <- function(x) {
  if (sum(!is.na(x)) < 2) return(NA_real_)
  sd(x, na.rm = TRUE)
}

safe_slope <- function(x, t) {
  x <- as.numeric(x)
  t <- as.numeric(t)
  
  keep <- is.finite(x) & is.finite(t)
  
  if (sum(keep) < 2) {
    return(NA_real_)
  }
  
  fit <- lm(x[keep] ~ t[keep])
  unname(coef(fit)[2])
}

safe_auc <- function(x, t) {
  keep <- complete.cases(x, t)
  if (sum(keep) < 2) return(NA_real_)
  
  x <- x[keep]
  t <- t[keep]
  
  sum(diff(t) * (head(x, -1) + tail(x, -1)) / 2)
}

safe_rel_change <- function(start, end) {
  if (is.na(start) || is.na(end) || start == 0) return(NA_real_)
  (end - start) / start
}

safe_cv <- function(x) {
  m <- mean(x, na.rm = TRUE)
  s <- safe_sd(x)
  if (is.na(m) || is.na(s) || m == 0) return(NA_real_)
  s / m
}

safe_mean <- function(x) {
  x <- as.numeric(x)
  x <- x[is.finite(x)]
  if (length(x) == 0) return(NA_real_)
  mean(x)
}

safe_time_to_thresh <- function(x, t, thresh, n_consecutive = 1) {
  pos <- !is.na(x) & x > thresh
  
  if (n_consecutive == 1) {
    if (!any(pos)) return(Inf)
    return(min(t[pos], na.rm = TRUE))
  }
  
  run_hit <- pos
  for (k in 1:(n_consecutive - 1)) {
    run_hit <- run_hit & lag(pos, k, default = FALSE)
  }
  
  if (!any(run_hit, na.rm = TRUE)) return(Inf)
  
  first_hit_idx <- which(run_hit)[1]
  start_idx <- first_hit_idx - (n_consecutive - 1)
  t[start_idx]
}

zscore_feature <- function(x) {
  mu <- mean(x, na.rm = TRUE)
  s  <- sd(x, na.rm = TRUE)
  
  if (is.na(s) || s == 0) {
    return(rep(NA_real_, length(x)))
  }
  
  (x - mu) / s
}
```

```{r}

spots_table = my_read_table("/Users/talfis/Desktop/LiveImaging/test/CompleteImage/A2_Merged_spots.csv")
#spots_table = my_read_table("/Users/talfis/Desktop/LiveImaging/test/CompleteImage/A3_Merged_spots.csv")

#Group and order by time
spots_table = spots_table %>%
  filter(!is.na(`Track.ID`))
spots_table = spots_table %>%
  arrange(`Track.ID`, `T..sec.`)

#Calculate speed
spots_table <- spots_table %>%
  group_by(`Track.ID`) %>%
  mutate(
    dX = X - lag(X),
    dY = Y - lag(Y),
    dT = `T..sec.` - lag(`T..sec.`),
    step_distance = sqrt(dX^2 + dY^2),
    speed = step_distance / dT
  ) %>%
  ungroup()

spots_table <- spots_table %>%
  mutate(speed = ifelse(is.na(speed), 0, speed))

spots_table <- spots_table %>%
  mutate(speed = ifelse(is.infinite(speed) | dT <= 0, NA, speed))

```

#Filtering split tracks (start at high GFP)
```{r}

time_to_green <- spots_table %>%
  group_by(Track.ID) %>%
  arrange(`T..sec.`, .by_group = TRUE) %>%
  summarise(
    time_to_green = first(`T..sec.`),
    first_gfp = first(Mean.ch2),
    .groups = "drop"
  )

ggplot(time_to_green, aes(x = time_to_green)) +
  geom_histogram(binwidth = 500) +
  labs(x = "Time to appearance (sec)", y = "Number of cells") +
  theme_bw()

time_to_green_1 = time_to_green[time_to_green$time_to_green<50000,]
time_to_green_2 = time_to_green[time_to_green$time_to_green>=100000,]

time_to_green_1$group <- "Early (<50000)"
time_to_green_2$group <- "Late (>=100000)"

combined_df <- bind_rows(time_to_green_1, time_to_green_2)
ggplot(combined_df, aes(x = first_gfp, color = group, fill = group)) +
  geom_density(alpha = 0.3) +
  labs(x = "GFP at first appearance", y = "Density") +
  theme_bw()

filtered_tracks <- time_to_green %>%
  filter(first_gfp < 2.5) %>%
  pull(Track.ID)

spots_table <- spots_table %>%
  filter(Track.ID %in% filtered_tracks)

length(unique(spots_table$Track.ID))
```

#Assign nuclei - simple nearest nucleus + temporal consistency
```{r}

nuc_table = my_read_table("/Users/talfis/Desktop/LiveImaging/test/CompleteImage/A2_nuclei_spots.csv")
#nuc_table = my_read_table("/Users/talfis/Desktop/LiveImaging/test/CompleteImage/A3_nuclei_spots.csv")

#----------------------------
# 0. clean tables
#----------------------------
spots_table <- spots_table %>%
  mutate(
    X = as.numeric(as.character(X)),
    Y = as.numeric(as.character(Y))
  ) %>%
  arrange(`Track.ID`, `T..sec.`)

nuc_table <- nuc_table %>%
  mutate(
    X = as.numeric(as.character(X)),
    Y = as.numeric(as.character(Y))
  ) %>%
  arrange(`Track.ID`, `T..sec.`)

#----------------------------
# 1. frame-wise nearest matches
#----------------------------
make_frame_matches_nearest <- function(t_now, nuc_table, spots_table, max_dist = 100) {
  
  nuc_sub <- nuc_table %>%
    filter(`T..sec.` == t_now) %>%
    transmute(
      nuc_track = `Track.ID`,
      t = `T..sec.`,
      nuc_x = X,
      nuc_y = Y
    )
  
  cell_sub <- spots_table %>%
    filter(`T..sec.` == t_now) %>%
    transmute(
      cell_track = `Track.ID`,
      t = `T..sec.`,
      cell_x = X,
      cell_y = Y
    )
  
  if (nrow(nuc_sub) == 0 || nrow(cell_sub) == 0) {
    return(NULL)
  }
  
  all_pairs <- merge(cell_sub, nuc_sub, by = "t") %>%
    mutate(
      centroid_dist = sqrt((cell_x - nuc_x)^2 + (cell_y - nuc_y)^2)
    ) %>%
    filter(centroid_dist <= max_dist)
  
  if (nrow(all_pairs) == 0) {
    return(NULL)
  }
  
  # mark the nearest nucleus for each cell in this frame
  all_pairs %>%
    group_by(cell_track, t) %>%
    mutate(
      is_nearest_for_cell = centroid_dist == min(centroid_dist, na.rm = TRUE)
    ) %>%
    ungroup()
}

all_times <- intersect(unique(nuc_table$`T..sec.`), unique(spots_table$`T..sec.`))

frame_matches <- map_dfr(
  all_times,
  make_frame_matches_nearest,
  nuc_table = nuc_table,
  spots_table = spots_table,
  max_dist = 100
)

#----------------------------
# 2. checkpoint setup
#----------------------------
checkpoint_props <- seq(0, 1, by = 0.1)

cell_track_times <- spots_table %>%
  group_by(cell_track = `Track.ID`) %>%
  summarise(
    t_min = min(`T..sec.`, na.rm = TRUE),
    t_max = max(`T..sec.`, na.rm = TRUE),
    .groups = "drop"
  )

checkpoint_targets <- cell_track_times %>%
  crossing(check_prop = checkpoint_props) %>%
  mutate(
    target_t = t_min + check_prop * (t_max - t_min),
    checkpoint = paste0("cp_", sprintf("%02d", round(check_prop * 100)))
  )

#----------------------------
# 3. checkpoint-level consistency
#----------------------------
checkpoint_scores <- checkpoint_targets %>%
  inner_join(frame_matches, by = "cell_track") %>%
  mutate(dt = abs(t - target_t)) %>%
  group_by(cell_track, nuc_track, checkpoint) %>%
  slice_min(dt, n = 1, with_ties = FALSE) %>%
  ungroup()

pair_checkpoint_summary <- checkpoint_scores %>%
  group_by(cell_track, nuc_track) %>%
  summarise(
    n_checkpoints_found = n(),
    cp_nearest_count = sum(is_nearest_for_cell, na.rm = TRUE),
    cp_frac_nearest = mean(is_nearest_for_cell, na.rm = TRUE),
    cp_mean_dist = mean(centroid_dist, na.rm = TRUE),
    cp_median_dist = median(centroid_dist, na.rm = TRUE),
    .groups = "drop"
  )

#----------------------------
# 4. full movie consistency
#----------------------------
pair_all_frame_summary <- frame_matches %>%
  group_by(cell_track, nuc_track) %>%
  summarise(
    n_overlap_frames = n(),
    nearest_count_all = sum(is_nearest_for_cell, na.rm = TRUE),
    frac_nearest_all = mean(is_nearest_for_cell, na.rm = TRUE),
    mean_dist_all = mean(centroid_dist, na.rm = TRUE),
    median_dist_all = median(centroid_dist, na.rm = TRUE),
    max_dist_all = max(centroid_dist, na.rm = TRUE),
    t_start = min(t, na.rm = TRUE),
    t_end = max(t, na.rm = TRUE),
    .groups = "drop"
  )

pair_summary <- pair_all_frame_summary %>%
  left_join(pair_checkpoint_summary, by = c("cell_track", "nuc_track")) %>%
  mutate(
    n_checkpoints_found = coalesce(n_checkpoints_found, 0L),
    cp_nearest_count = coalesce(cp_nearest_count, 0),
    cp_frac_nearest = coalesce(cp_frac_nearest, 0),
    cp_mean_dist = coalesce(cp_mean_dist, Inf),
    cp_median_dist = coalesce(cp_median_dist, Inf)
  )

#----------------------------
# 5. choose best nucleus for each cell
#----------------------------
best_nucleus_per_cell <- pair_summary %>%
  group_by(cell_track) %>%
  arrange(
    desc(cp_nearest_count),
    desc(cp_frac_nearest),
    desc(frac_nearest_all),
    cp_mean_dist,
    mean_dist_all,
    .by_group = TRUE
  ) %>%
  slice(1) %>%
  ungroup()

#----------------------------
# 6. resolve nuclei assigned to several cells
#----------------------------
best_unique_assignments <- best_nucleus_per_cell %>%
  group_by(nuc_track) %>%
  arrange(
    desc(cp_nearest_count),
    desc(cp_frac_nearest),
    desc(frac_nearest_all),
    cp_mean_dist,
    mean_dist_all,
    .by_group = TRUE
  ) %>%
  slice(1) %>%
  ungroup()

#----------------------------
# 7. confidence and filtering
#----------------------------
final_assignments <- best_unique_assignments %>%
  mutate(
    assignment_confidence = case_when(
      n_overlap_frames >= 5 &
        frac_nearest_all >= 0.8 &
        cp_frac_nearest >= 0.8 ~ "high",
      
      n_overlap_frames >= 5 &
        frac_nearest_all >= 0.6 &
        cp_frac_nearest >= 0.6 ~ "medium",
      
      TRUE ~ "low"
    )
  )

good_assignments <- final_assignments %>%
  filter(
    n_overlap_frames >= 5,
    frac_nearest_all >= 0.6,
    cp_frac_nearest >= 0.6
  )

#----------------------------
# 8. QC
#----------------------------
cell_conflicts_after_resolution <- good_assignments %>%
  count(cell_track) %>%
  filter(n > 1)

nuc_conflicts_after_resolution <- good_assignments %>%
  count(nuc_track) %>%
  filter(n > 1)

one_to_one_ok <- nrow(good_assignments) == n_distinct(good_assignments$cell_track) &&
  nrow(good_assignments) == n_distinct(good_assignments$nuc_track)

#----------------------------
# 9. relabel nucleus table into cell IDs
#----------------------------
nuc_table_filt <- nuc_table %>%
  inner_join(
    good_assignments %>% select(cell_track, nuc_track),
    by = c("Track.ID" = "nuc_track")
  ) %>%
  mutate(`Track.ID` = cell_track) %>%
  select(-cell_track)

# optional summaries
assignment_summary <- tibble(
  n_total_cells = n_distinct(spots_table$`Track.ID`),
  n_cells_kept = n_distinct(good_assignments$cell_track),
  pct_cells_kept = 100 * n_cells_kept / n_total_cells
)

length(unique(nuc_table_filt$Track.ID))
```

#BFP -> GFP bleedthrough correction
```{r}
# Estimate bleedthrough coefficient alpha from first n_baseline frames of each
# track. After filtering (first_gfp < 2.5), these frames are at or before true
# GFP onset, so observed ch2 ≈ alpha * ch1 (bleedthrough only).

n_baseline <- 5

baseline_df <- spots_table %>%
  filter(Track.ID %in% filtered_tracks) %>%
  arrange(Track.ID, `T..sec.`) %>%
  group_by(Track.ID) %>%
  slice_head(n = n_baseline) %>%
  ungroup() %>%
  filter(is.finite(Mean.ch1), Mean.ch1 > 0, is.finite(Mean.ch2))

# Global fit
bleed_fit0 <- lm(Mean.ch2 ~ Mean.ch1 - 1, data = baseline_df)   # through origin
bleed_fit1 <- lm(Mean.ch2 ~ Mean.ch1,     data = baseline_df)   # with intercept

bleed_alpha     <- unname(coef(bleed_fit0)[1])
bleed_alpha_int <- unname(coef(bleed_fit1)[2])
bleed_intercept <- unname(coef(bleed_fit1)[1])

cat(sprintf("alpha (through origin):  %.5f  R2 = %.4f\n",
            bleed_alpha, summary(bleed_fit0)$r.squared))
cat(sprintf("alpha (with intercept):  %.5f  R2 = %.4f  intercept = %.4f\n",
            bleed_alpha_int, summary(bleed_fit1)$r.squared, bleed_intercept))

# Per-cell alpha to assess consistency
per_cell_alpha <- baseline_df %>%
  group_by(Track.ID) %>%
  summarise(
    n = n(),
    alpha_cell = if (n() >= 3 && sd(Mean.ch1, na.rm = TRUE) > 1e-6)
      unname(coef(lm(Mean.ch2 ~ Mean.ch1 - 1))[1])
    else NA_real_,
    .groups = "drop"
  )

# Diagnostic plots
p_scatter <- ggplot(baseline_df, aes(x = Mean.ch1, y = Mean.ch2)) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_abline(slope = bleed_alpha,     intercept = 0,               color = "red",  linewidth = 1) +
  geom_abline(slope = bleed_alpha_int, intercept = bleed_intercept, color = "blue", linewidth = 1, linetype = "dashed") +
  labs(
    title = sprintf("BFP → GFP bleedthrough  (first %d frames per cell)", n_baseline),
    subtitle = sprintf("Through origin: α=%.4f  R²=%.3f   |   With intercept: α=%.4f  R²=%.3f",
                       bleed_alpha,     summary(bleed_fit0)$r.squared,
                       bleed_alpha_int, summary(bleed_fit1)$r.squared),
    x = "Cell BFP (ch1)", y = "Cell GFP (ch2)"
  ) + theme_bw()

p_alpha_dist <- ggplot(per_cell_alpha %>% filter(is.finite(alpha_cell)),
                       aes(x = alpha_cell)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  geom_vline(xintercept = bleed_alpha, color = "red",    linewidth = 1) +
  geom_vline(xintercept = median(per_cell_alpha$alpha_cell, na.rm = TRUE),
             color = "orange", linewidth = 1, linetype = "dashed") +
  labs(title = "Per-cell bleedthrough coefficient α",
       x = "α", y = "Cells") + theme_bw()

print(p_scatter)
print(p_alpha_dist)

# Apply correction (use through-origin alpha; swap to bleed_alpha_int if preferred)
spots_table <- spots_table %>%
  mutate(Mean.ch2.corrected = Mean.ch2 - bleed_alpha * Mean.ch1)
```

#Red rising rate
```{r}
red_thresh_cell <- 2.25

spots <- spots_table %>%
  arrange(`Track.ID`, Frame) %>%
  group_by(`Track.ID`) %>%
  mutate(
    ch3_pos = Mean.ch3 > red_thresh_cell,
    red_now = ch3_pos &
      lag(ch3_pos, 1, default = FALSE) &
      lag(ch3_pos, 2, default = FALSE) &
      lag(ch3_pos, 3, default = FALSE) &
      lag(ch3_pos, 4, default = FALSE)
  ) %>%
  ungroup()

spots <- spots %>%
  group_by(`Track.ID`) %>%
  mutate(
    red_onset_candidate = ifelse(
      any(red_now, na.rm = TRUE),
      min(Frame[red_now], na.rm = TRUE) - 4,
      NA_real_
    )
  ) %>%
  ungroup()

red_rate_qc <- spots %>%
  filter(!is.na(red_onset_candidate)) %>%
  group_by(`Track.ID`) %>%
  filter(
    Frame >= first(red_onset_candidate),
    Frame <= first(red_onset_candidate) + 4
  ) %>%
  summarise(
    red_rise_rate = {
      t <- `T..sec.` - min(`T..sec.`, na.rm = TRUE)
      y <- Mean.ch3
      keep <- is.finite(t) & is.finite(y)

      if (sum(keep) >= 3) {
        unname(coef(lm(y[keep] ~ t[keep]))[2])
      } else {
        NA_real_
      }
    },
    red_int_start = first(Mean.ch3),
    red_int_end   = last(Mean.ch3),
    red_delta_5f  = red_int_end - red_int_start,
    .groups = "drop"
  )

red_rate_qc <- spots %>%
  left_join(red_rate_qc, by = "Track.ID")

slope_cutoff <- 0.002
delta_cutoff <- 0.3 

red_rate_qc <- red_rate_qc %>%
  mutate(
    red_start_valid = !is.na(red_onset_candidate) &
      !is.na(red_rise_rate) &
      red_rise_rate > slope_cutoff &
      red_delta_5f > delta_cutoff
  )


red_start_summary <- red_rate_qc %>%
  group_by(`Track.ID`) %>%
  summarise(
    red_start_frame = ifelse(any(red_start_valid), first(red_onset_candidate), NA_real_),
    red_rise_rate = first(red_rise_rate),
    red_delta_5f = first(red_delta_5f),
    .groups = "drop"
  )

red_rate_qc = red_rate_qc[red_rate_qc$red_rise_rate>0 & red_rate_qc$red_delta_5f>0,]


ggplot(red_rate_qc, aes(x = red_rise_rate*60)) +
  geom_histogram(bins = 40, fill = "red", alpha = 0.6) +
  theme_minimal() + labs(x = "mCherry signal rise per minute")

slope_cutoff <- quantile(red_rate_qc$red_rise_rate[red_rate_qc$red_rise_rate > 0], 0.1, na.rm = TRUE)
delta_cutoff <- quantile(red_rate_qc$red_delta_5f[red_rate_qc$red_delta_5f > 0], 0.1, na.rm = TRUE)



```


#Time to GFP and time to mCherry
```{r}

smooth_k <- 3
baseline_n <- 5

gfp_thresh <- 0
blue_thresh <- 0
blue_run <- 3
gfp_run <- 3

red_run <- 5
red_k_sd <- 3
rise_lag <- 3

spots2 <- spots_table %>%
  filter(Track.ID %in% unique(nuc_table_filt$Track.ID)) %>%
  arrange(Track.ID, `T..sec.`) %>%
  group_by(Track.ID) %>%
  mutate(
    first_time = first(`T..sec.`),
    norm_time = `T..sec.` - first_time,

    ch2_smooth = roll_mean_partial(Mean.ch2, smooth_k),
    ch3_smooth = roll_mean_partial(Mean.ch3, smooth_k),

    # per-cell early baseline
    ch3_baseline = mean(head(ch3_smooth, baseline_n), na.rm = TRUE),
    ch3_baseline_sd = sd(head(ch3_smooth, baseline_n), na.rm = TRUE),

    # upward change relative to recent past
    ch3_rise = ch3_smooth - lag(ch3_smooth, rise_lag),
    gfp_pos = !is.na(ch2_smooth) & ch2_smooth > gfp_thresh,

    red_candidate = !is.na(ch3_smooth) &
    ch3_smooth > red_thresh_cell,

    green_now = gfp_pos &
      lag(gfp_pos, 1, default = FALSE) &
      lag(gfp_pos, 2, default = FALSE),

    # candidate red call: 5 consecutive red_candidate frames
    red_now_candidate = red_candidate &
      lag(red_candidate, 1, default = FALSE) &
      lag(red_candidate, 2, default = FALSE) &
      lag(red_candidate, 3, default = FALSE) &
      lag(red_candidate, 4, default = FALSE)
  ) %>%
  mutate(
    green_start_time = ifelse(green_now, lag(`T..sec.`, gfp_run - 1), NA_real_),
    green_complete_time = ifelse(green_now, `T..sec.`, NA_real_),

    # first frame of the 5-frame red run
    red_start_time_candidate = ifelse(
      any(red_now_candidate, na.rm = TRUE),
      min(`T..sec.`[red_now_candidate], na.rm = TRUE) - 4 * median(diff(`T..sec.`), na.rm = TRUE),
      NA_real_
    ),

    red_complete_time_candidate = ifelse(
      any(red_now_candidate, na.rm = TRUE),
      min(`T..sec.`[red_now_candidate], na.rm = TRUE),
      NA_real_
    )
  ) %>%
  mutate(
    # local 5-frame window starting at candidate onset
    in_red_qc_window = !is.na(red_start_time_candidate) &
      `T..sec.` >= red_start_time_candidate &
      `T..sec.` <= red_start_time_candidate + 4 * median(diff(`T..sec.`), na.rm = TRUE)
  ) %>%
  mutate(
    red_rise_rate = {
      keep <- in_red_qc_window & is.finite(`T..sec.`) & is.finite(ch3_smooth)
      if (sum(keep) >= 3) {
        t <- `T..sec.`[keep] - min(`T..sec.`[keep], na.rm = TRUE)
        y <- ch3_smooth[keep]
        unname(coef(lm(y ~ t))[2])
      } else {
        NA_real_
      }
    },

    red_delta_5f = {
      keep <- in_red_qc_window & is.finite(ch3_smooth)
      if (sum(keep) >= 2) {
        last(ch3_smooth[keep]) - first(ch3_smooth[keep])
      } else {
        NA_real_
      }
    }
  ) %>%
  mutate(
    red_start_valid =
      !is.na(red_start_time_candidate) &
      !is.na(red_rise_rate) &
      !is.na(red_delta_5f) &
      red_rise_rate >= slope_cutoff &
      red_delta_5f >= delta_cutoff,

    # final validated red timings
    red_start_time = ifelse(red_start_valid, red_start_time_candidate, NA_real_),
    red_complete_time = ifelse(red_start_valid, red_complete_time_candidate, NA_real_)
  ) %>%
  mutate(
    # final red_now: all frames from validated onset onward
    red_now = !is.na(red_start_time) & `T..sec.` >= red_start_time
  ) %>%
  ungroup()


spots3 <- spots2 %>%
  rename_with(~ paste0("cell_", .x), -c(`Track.ID`, `T..sec.`, `Frame`)) %>%
  left_join(
    nuc_table_filt %>%
      rename_with(~ paste0("nuc_", .x), -c(`Track.ID`, `T..sec.`, `Frame`)),
    by = c("Track.ID", "Frame", "T..sec.")
  ) %>%
  arrange(`Track.ID`, Frame) %>%
  group_by(`Track.ID`) %>%
  mutate(
    # smooth channels
    blue_smooth = roll_mean_partial(nuc_Mean.ch1, smooth_k),
    green_smooth = roll_mean_partial(cell_Mean.ch2, smooth_k),
    red_smooth = roll_mean_partial(cell_Mean.ch3, smooth_k),

    # positive flags
    blue_pos = !is.na(blue_smooth) & blue_smooth > blue_thresh,
    green_pos = !is.na(green_smooth) & green_smooth > gfp_thresh,
    red_pos = !is.na(red_smooth) & red_smooth > red_thresh_cell,

    # consecutive-frame event completion flags
    blue_now = blue_pos &
      lag(blue_pos, 1, default = FALSE) &
      lag(blue_pos, 2, default = FALSE),

    green_now = green_pos &
      lag(green_pos, 1, default = FALSE) &
      lag(green_pos, 2, default = FALSE),

    # red_now: use validated onset from spots2 (5-consecutive + slope/delta QC)
    # loops through ALL valid 5-frame windows, not just the first one
    red_now = cell_red_now,

    # completion times
    blue_complete_time  = ifelse(blue_now, `T..sec.`, NA_real_),
    green_complete_time = ifelse(green_now, `T..sec.`, NA_real_),
    red_complete_time   = cell_red_complete_time,

    # start times of the run
    blue_start_time  = ifelse(blue_now, lag(`T..sec.`, blue_run - 1), NA_real_),
    green_start_time = ifelse(green_now, lag(`T..sec.`, gfp_run - 1), NA_real_),
    red_start_time   = cell_red_start_time
  ) %>%
  ungroup()


#Normalized cyto signal
 spots3 <- spots3 %>%
   mutate(
     A_cell = spots3$cell_Area,
     A_nuc  = spots3$nuc_Area,
     A_cyto = pmax(A_cell - A_nuc, 1),

     green_cell_total = cell_Mean.ch2 * A_cell,
     green_nuc_total  = nuc_Mean.ch2 * A_nuc,
     green_cyto_total = pmax(green_cell_total - green_nuc_total, 0),
     green_cyto_mean = green_cyto_total / A_cyto,
     valid_geom = A_nuc > 0 & A_cell > A_nuc) %>%
     filter(valid_geom)
   )

cor(spots3$green_cyto_mean, spots3$nuc_Mean.ch1, use = "complete.obs")   # blue vs corrected green
cor(spots3$nuc_Mean.ch2, spots3$nuc_Mean.ch1, use = "complete.obs")          # blue vs original green

time_summary <- spots3 %>%
  group_by(`Track.ID`) %>%
  summarise(
    first_track_time = min(`T..sec.`, na.rm = TRUE),

    time_to_blue_complete = ifelse(
      any(blue_now, na.rm = TRUE),
      min(blue_complete_time, na.rm = TRUE),
      Inf
    ),
    time_to_blue_start = ifelse(
      any(blue_now, na.rm = TRUE),
      min(blue_start_time, na.rm = TRUE),
      Inf
    ),

    time_to_green_complete = ifelse(
      any(green_now, na.rm = TRUE),
      min(green_complete_time, na.rm = TRUE),
      Inf
    ),
    time_to_green_start = ifelse(
      any(green_now, na.rm = TRUE),
      min(green_start_time, na.rm = TRUE),
      Inf
    ),

    time_to_red_complete = ifelse(
      any(red_now, na.rm = TRUE),
      min(red_complete_time, na.rm = TRUE),
      Inf
    ),
    time_to_red_start = ifelse(
      any(red_now, na.rm = TRUE),
      min(red_start_time, na.rm = TRUE),
      Inf
    ),

    delay_green_to_blue = ifelse(
      is.finite(time_to_blue_start) & is.finite(time_to_green_start),
      time_to_blue_start - time_to_green_start,
      Inf
    ),

    delay_green_to_red = ifelse(
      is.finite(time_to_green_start) & is.finite(time_to_red_start),
      time_to_red_start - time_to_green_start,
      Inf
    ),
    
    green_at_red_start = ifelse(
      any(red_now, na.rm = TRUE),
      cell_Mean.ch2[which(red_now)[1]],
      NA_real_
    ),
    
    green_one_frame_before_red_start = ifelse(
      any(red_now, na.rm = TRUE) && which(red_now)[1] > 1,
      cell_Mean.ch2[which(red_now)[1] - 1],
      NA_real_
    ),

    first_blue_smooth = first(blue_smooth[!is.na(blue_smooth)]),
    first_green_smooth = first(green_smooth[!is.na(green_smooth)]),
    first_red_smooth = first(red_smooth[!is.na(red_smooth)]),

    max_blue = max(nuc_Mean.ch1, na.rm = TRUE),
    max_gfp = max(cell_Mean.ch2, na.rm = TRUE),
    max_red = max(cell_Mean.ch3, na.rm = TRUE),

    .groups = "drop"
  )


min_plausible_delay_min <- 240


time_summary <- time_summary %>%
  mutate(
    time_to_green_start_min = ifelse(is.finite(time_to_green_start), time_to_green_start / 60, NA),
    time_to_red_start_min = ifelse(is.finite(time_to_red_start), time_to_red_start / 60, NA),
    delay_green_to_red_min = ifelse(is.finite(delay_green_to_red), delay_green_to_red / 60, NA),

    suspicious_short_delay = !is.na(delay_green_to_red_min) & delay_green_to_red_min < min_plausible_delay_min
  )

short_delay_cells <- time_summary %>%
  filter(suspicious_short_delay) %>%
  arrange(delay_green_to_red_min)

time_summary %>%
  summarise(
    n_cells = n(),
    n_green = sum(is.finite(time_to_green_start)),
    n_red = sum(is.finite(time_to_red_start)),
    n_short_delay = sum(suspicious_short_delay, na.rm = TRUE)
  )

suspect_ids <- short_delay_cells$Track.ID
#remove more tracks
suspect_ids = c(suspect_ids, 9)

time_summary = time_summary %>%
  filter(!Track.ID %in% suspect_ids)

spots2 = spots2 %>%
  filter(!Track.ID %in% suspect_ids)

nuc_table_filt = nuc_table_filt %>%
  filter(!Track.ID %in% suspect_ids)

```

```{r}
spots_table_tmp = spots3

spots3 <- spots3 %>%
  filter(Track.ID %in% unique(nuc_table_filt$Track.ID))


time_to_red_plot <- time_summary %>%
  filter(is.finite(time_to_red_complete)) %>%
  arrange(time_to_red_complete) %>%
  mutate(
    cell_order = row_number(),
    time_to_red_complete = time_to_red_complete / 60
  )

time_summary_flat = time_summary
rownames(time_summary_flat) = time_summary_flat$Track.ID
time_summary_flat = time_summary_flat[,c("time_to_green_start", "time_to_red_complete", "delay_green_to_red")]
time_summary_flat = time_summary_flat %>%
  arrange(time_to_green_start) %>%
  mutate(
    cell_order = row_number(),
    time_to_red = time_to_red_complete / 60,
    time_to_gfp = time_to_green_start / 60,
    time_green_to_red = delay_green_to_red / 60
  )
time_summary_flat[is.infinite(time_summary_flat$time_to_red),"time_to_red"] = NA

ggplot(time_summary_flat, aes(y = cell_order)) +
  geom_segment(aes(x = 0, xend = time_to_gfp, yend = cell_order), alpha = 0.4) +
  geom_point(aes(x = time_to_gfp), size = 1.8) +
  labs(title = "GFP start", x = "Time since start (minutes)", y = "Cells") + theme_bw() + xlim(0,4000)

ggplot(time_to_red_plot, aes(y = cell_order)) +
  geom_segment(aes(x = 0, xend = time_to_red_complete, yend = cell_order), alpha = 0.4) +
  geom_point(aes(x = time_to_red_complete), size = 1.8) +
  labs(title = "mCherry start", x = "Time since start (minutes)", y = "Cells") + theme_bw() + xlim(0,4000)

ggplot(time_to_red_plot, aes(y = cell_order)) +
  geom_segment(aes(x = 0, xend = delay_green_to_red_min , yend = cell_order), alpha = 0.4) +
  geom_point(aes(x = delay_green_to_red_min), size = 1.8) +
  labs(title = "mCherry start",x = "Time since GFP (minutes)", y = "Cells") + theme_bw() + xlim(0,4000)


ggplot(time_summary_flat, aes(y = cell_order)) +
  geom_segment(aes(x = 0, xend = time_to_gfp, yend = cell_order), alpha = 0.1) +
  geom_point(aes(x = time_to_gfp), size = 1.8, color = "green") +
  geom_segment(aes(x = 0, xend = time_to_red, yend = cell_order), alpha = 0.05) +
  geom_point(aes(x = time_to_red), size = 1.8, color = "red") +
  labs(x = "Time since start (minutes)", y = "Cells") + theme_bw() + xlim(0,4000)


ggplot() + 
  geom_histogram(data = time_summary_flat, aes(x = time_to_gfp), fill = "green", alpha = 0.4, bins = 200) + 
  geom_histogram(data = time_summary_flat, aes(x = time_to_red), fill = "red", alpha = 0.4, bins = 200) + xlim(0,4000) + labs(x = "Time (minutes)")


time_summary$turned_red = "yes"
time_summary[is.infinite(time_summary$time_to_red_complete),"turned_red"] = "no"
  
ggplot(time_summary, aes(x = turned_red)) + geom_bar(width = 0.5) + theme_bw()


time_summary_plot <- time_summary %>%
  arrange(time_to_green_start) %>%
  mutate(
    cell_order = row_number(),
    time_to_blue_start = ifelse(is.infinite(time_to_blue_start), NA, time_to_blue_start / 60),
    time_to_green_start = ifelse(is.infinite(time_to_green_start), NA, time_to_green_start / 60),
    time_to_red_start = ifelse(is.infinite(time_to_red_start), NA, time_to_red_start / 60),
    delay_green_to_red_min = ifelse(is.finite(delay_green_to_red), delay_green_to_red / 60, NA),
    delay_green_to_blue_min = ifelse(is.finite(delay_green_to_blue), delay_green_to_blue / 60, NA),
    suspicious_short_delay = !is.na(delay_green_to_red_min) & delay_green_to_red_min < min_plausible_delay_min
  )

time_summary_plot = time_summary_plot[time_summary_plot$suspicious_short_delay == FALSE,]

ggplot(time_summary_plot, aes(y = cell_order)) +
  
  geom_segment(aes(x = 0, xend = time_to_green_start, yend = cell_order), alpha = 0.1, color = "green") +
  geom_point(aes(x = time_to_green_start), size = 1.5, color = "green") +
  
  geom_segment(aes(x = 0, xend = time_to_blue_start, yend = cell_order), alpha = 0.1, color = "blue") +
  geom_point(aes(x = time_to_blue_start), size = 1.5, color = "blue") +

  geom_segment(aes(x = 0, xend = time_to_red_start, yend = cell_order), alpha = 0.25, color = "red") +
  geom_point(aes(x = time_to_red_start), size = 1.5, color = "red") +

  labs(
    x = "Time since start of movie (minutes)",
    y = "Cells",
    title = "Blue, green and red first appearance"
  ) +
  theme_bw() +
  coord_cartesian(xlim = c(0, 4000)) + coord_flip()

ggplot() + 
  geom_histogram(data = time_summary_plot, aes(x = time_to_green_start), fill = "green", alpha = 0.4, bins = 200) + 
  geom_histogram(data = time_summary_plot, aes(x = time_to_blue_start), fill = "blue", alpha = 0.4, bins = 200) + 
  geom_histogram(data = time_summary_plot, aes(x = time_to_red_start), fill = "red", alpha = 0.4, bins = 200) + xlim(0,4000) + labs(x = "Time (minutes)")

ggplot() + 
  geom_density(data = time_summary_plot, aes(x = time_to_green_start), fill = "green", alpha = 0.4) + 
  geom_density(data = time_summary_plot, aes(x = time_to_blue_start), fill = "blue", alpha = 0.4) + 
  geom_density(data = time_summary_plot, aes(x = time_to_red_start), fill = "red", alpha = 0.4) + xlim(0,4000) + labs(x = "Time (minutes)")

ggplot(time_summary_plot, aes(x = delay_green_to_red_min)) + geom_density(fill = "red") + labs(title = "green to red delay (minutes)")

gfp_at_red <- spots3 %>%
  arrange(`Track.ID`, Frame) %>%
  group_by(`Track.ID`) %>%
  filter(is.finite(red_start_time)) %>%                
  slice(1) %>%                      
  ungroup()



ggplot(time_summary, aes(x = green_at_red_start)) + geom_density(fill = "green") + labs(title = "green intensity at red transition")
ggplot(time_summary, aes(x = green_one_frame_before_red_start)) + geom_density(fill = "green") + labs(title = "green intensity one frame before red transition")

draw_it(time_summary, "green_at_red_start", "delay_green_to_red")
draw_it(time_summary, "green_one_frame_before_red_start", "delay_green_to_red")


```
#Green threshold for red transition and GFP distribution for early infected cells
```{r}

spots_thresh <- spots3 %>%
  arrange(`Track.ID`, Frame) %>%
  group_by(`Track.ID`) %>%
  mutate(
    first_red_frame = ifelse(
      any(red_now),
      min(Frame[red_now]),
      Inf
    )
  )

spots_thresh <- spots_thresh %>%
  mutate(
    red_in_next_3 = 
      lead(red_now, 1, default = FALSE) |
      lead(red_now, 2, default = FALSE) |
      lead(red_now, 3, default = FALSE)
  ) %>%
  ungroup()

spots_thresh <- spots_thresh %>%
  filter(Frame < first_red_frame)



ggplot(spots_thresh, aes(x = cell_Mean.ch2, fill = red_in_next_3)) +
  geom_density(alpha = 0.4) +
  theme_minimal()

spots3 <- spots3 %>%
  group_by(`Track.ID`) %>%
  mutate(norm_frame = Frame - min(Frame)) %>%
  ungroup()


ggplot(time_summary_plot, aes(x = delay_green_to_red_min)) +
  geom_density(alpha = 0.4, fill = "Green") +
  theme_minimal() + labs(title = "green to red time")


ggplot(time_summary_plot[time_summary_plot$time_to_green_start_min<1000,], aes(x = delay_green_to_red_min)) +
  geom_density(alpha = 0.4, fill = "Green") +
  theme_minimal() + labs(title = "green to red time for early infected")

early_g2r <- time_summary_plot %>%
  filter(delay_green_to_red_min < 1000) %>%
  pull(Track.ID) %>%
  na.omit()

```

#Individual tracks
```{r}

spots3_early_g2r = spots3 %>%
  filter(Track.ID %in% early_g2r)

p_GFP = plot_metric(spots3, "cell_Mean.ch2", "Mean intensity ch2", "Track intensity over time: GFP")
p_GFPcyto = plot_metric(spots3, "green_cyto_mean", "Mean intensity ch2", "Track intensity over time: GFP") + ylim(0,100)
p_BFP = plot_metric(spots3, "nuc_Mean.ch1", "Mean intensity ch1", "Track intensity over time: BFP")
p_mCherry = plot_metric(spots3, "cell_Mean.ch3", "Mean intensity ch3", "Track intensity over time: mCherry")
p_area = plot_metric(spots3, "cell_Area",     "Cell size",               "cell size over time")
p_nuc_area = plot_metric(spots3, "cell_Area", "nuc size", "nuc size over time: BFP")
p_speed = plot_metric(spots3, "cell_speed",    "Speed",              "Track speed over time")

print(p_GFP)
print(p_GFPcyto)
print(p_BFP)
print(p_mCherry)
print(p_area)
print(p_nuc_area)
print(p_speed)

p_GFP = plot_metric(spots3_early_g2r, "cell_Mean.ch2", "Mean intensity ch2", "Track intensity over time: GFP")
p_mCherry = plot_metric(spots3_early_g2r, "cell_Mean.ch3", "Mean intensity ch3", "Track intensity over time: mCherry") + ylim(1,5)
print(p_GFP)
print(p_mCherry)


red_times <- time_summary_plot %>%
  select(Track.ID, time_to_red_start) 

mcherry_tracks <- spots3_early_g2r %>%
  filter(!is.na(cell_Mean.ch3)) %>%
  distinct(Track.ID) %>%
  arrange(Track.ID) %>%
  mutate(page = ceiling(row_number() / 10))

spots3_mcherry_pages <- spots3_early_g2r %>%
  left_join(mcherry_tracks, by = "Track.ID")

spots3_mcherry_pages <- spots3_mcherry_pages %>%
  left_join(red_times, by = "Track.ID")

for (p in sort(unique(spots3_mcherry_pages$page))) {
  print(
    spots3_mcherry_pages %>%
      filter(page == p) %>%
      ggplot(aes(x = T..sec.  / 60, y = cell_Mean.ch3)) +
      geom_line() +
      geom_point(size = 0.5) +
      
      geom_vline(
        aes(xintercept = time_to_red_start),
        color = "red",
        linetype = "dotted",
        linewidth = 0.7,
        na.rm = TRUE
      ) +
      
      facet_wrap(~ Track.ID, scales = "free_y", ncol = 2) +
      labs(
        x = "Time from track start (min)",
        y = "mCherry intensity",
        title = paste("mCherry tracks - page", p)
      ) + ylim(1.5,5) +
      theme_bw()
  )
}

```

```{r}
# Population summary per timepoint


time_summary_merge <- spots2 %>%
  rename_with(~ paste0("cell_", .x), -c(`Track.ID`, `T..sec.`)) %>%
  full_join(
    nuc_table_filt %>%
      rename_with(~ paste0("nuc_", .x), -c(`Track.ID`, `T..sec.`)),
    by = c("Track.ID", "T..sec.")
  ) %>%
  arrange(`Track.ID`, `T..sec.`)

time_summary_roll = time_summary_merge

time_summary_roll <- time_summary_roll %>%
  group_by(T..sec.) %>%
  summarise(
    mean_ch1 = mean(nuc_Mean.ch1, na.rm = TRUE),
    mean_ch2 = mean(cell_Mean.ch2, na.rm = TRUE),
    mean_ch3 = mean(cell_Mean.ch3, na.rm = TRUE),
    mean_area = mean(cell_Area, na.rm = TRUE),
    mean_nuc_area = mean(nuc_Area, na.rm = TRUE),
    mean_nuc_circ = mean(nuc_Circ., na.rm = TRUE),
    mean_speed = mean(cell_speed, na.rm = TRUE),

    sem_ch1 = sd(nuc_Mean.ch1, na.rm = TRUE) / sqrt(n()),
    sem_ch2 = sd(cell_Mean.ch2, na.rm = TRUE) / sqrt(n()),
    sem_ch3 = sd(cell_Mean.ch3, na.rm = TRUE) / sqrt(n()),
    sem_nuc_area = sd(nuc_Area, na.rm = TRUE) / sqrt(n()),
    sem_nuc_circ = sd(nuc_Circ., na.rm = TRUE) / sqrt(n()),
    sem_area = sd(cell_Area, na.rm = TRUE) / sqrt(n()),
    sem_speed = sd(cell_speed, na.rm = TRUE) / sqrt(n()),

    n_cells = n(),
    .groups = "drop"
  ) %>%
  arrange(T..sec.)

# Sliding SD window

window_size <- 5

time_summary_roll <- time_summary_roll %>%
  mutate(
    roll_sem_ch1 = rollapply(sem_ch1, window_size, mean, fill = NA, align = "center"),
    roll_sem_ch2 = rollapply(sem_ch2, window_size, mean, fill = NA, align = "center"),
    roll_sem_ch3 = rollapply(sem_ch3, window_size, mean, fill = NA, align = "center"),
    roll_sem_area = rollapply(sem_area, window_size, mean, fill = NA, align = "center"),
    roll_sem_nuc_area = rollapply(sem_nuc_area, window_size, mean, fill = NA, align = "center"),
    roll_sem_nuc_circ = rollapply(sem_nuc_circ, window_size, mean, fill = NA, align = "center"),
    roll_sem_speed = rollapply(sem_speed, window_size, mean, fill = NA, align = "center")
  )


p1 <- plot_mean_sd(time_summary_roll, "mean_ch1", "roll_sem_ch1",
                   "Mean intensity BFP",
                   "Sliding mean: BFP")

p2 <- plot_mean_sd(time_summary_roll, "mean_ch2", "roll_sem_ch2",
                   "Mean intensity GFP",
                   "Sliding mean: GFP")

p3 <- plot_mean_sd(time_summary_roll, "mean_ch3", "roll_sem_ch3",
                   "Mean intensity mCherry",
                   "Sliding mean: mCherry")

p4 <- plot_mean_sd(time_summary_roll, "mean_area", "roll_sem_area",
                   "Cell size",
                   "Sliding mean: cell area")

p5 <- plot_mean_sd(time_summary_roll, "mean_speed", "roll_sem_speed",
                   "Cell speed",
                   "Sliding mean: cell speed")

p6 <- plot_mean_sd(time_summary_roll, "mean_nuc_area", "roll_sem_nuc_area",
                   "Nuclei size",
                   "Sliding mean: Nuclei size")

p7 <- plot_mean_sd(time_summary_roll, "mean_nuc_circ", "roll_sem_nuc_circ",
                   "Nuclei circ",
                   "Sliding mean: Nuclei circ")

print(p1)
print(p2)
print(p3)
print(p4)
print(p5)
print(p6)
print(p7)

# Inspect how many cells contribute

ggplot(time_summary_roll, aes(x = T..sec., y = n_cells)) +
  geom_line() +
  geom_point() +
  theme_bw() +
  labs(
    title = "Cells contributing per timepoint",
    x = "Time (sec)",
    y = "Cell count"
  )

time_summary_roll[is.na(time_summary_roll)] = 0

time_summary_roll <- time_summary_roll %>%
  arrange(T..sec.) %>%
  mutate(
    roll_mean_ch1 = rollapply(
      mean_ch1, window_size,
      function(x) if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE),
      fill = NA, align = "center", partial = TRUE
    ),
    roll_mean_ch2 = rollapply(
      mean_ch2, window_size,
      function(x) if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE),
      fill = NA, align = "center", partial = TRUE
    ),
    roll_mean_ch3 = rollapply(
      mean_ch3, window_size,
      function(x) if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE),
      fill = NA, align = "center", partial = TRUE
    ),

    baseline_ch1 = ifelse(any(mean_ch1 > 0, na.rm = TRUE),
                          first(mean_ch1[mean_ch1 > 0]),
                          0),

    roll_mean_ch1 = roll_mean_ch1 - baseline_ch1,
    roll_mean_ch1 = pmax(roll_mean_ch1, 0),

    roll_mean_ch2 = roll_mean_ch2 - first(roll_mean_ch2[!is.na(roll_mean_ch2)]),
    roll_mean_ch3 = roll_mean_ch3 - first(roll_mean_ch3[!is.na(roll_mean_ch3)]),

    roll_sem_ch1 = rollapply(
      sem_ch1, window_size,
      function(x) if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE),
      fill = NA, align = "center", partial = TRUE
    ),
    roll_sem_ch2 = rollapply(
      sem_ch2, window_size,
      function(x) if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE),
      fill = NA, align = "center", partial = TRUE
    ),
    roll_sem_ch3 = rollapply(
      sem_ch3, window_size,
      function(x) if (all(is.na(x))) NA_real_ else mean(x, na.rm = TRUE),
      fill = NA, align = "center", partial = TRUE
    ),

    time_min = T..sec. / 60
  ) %>%
  select(-baseline_ch1)


# convert to long format

plot_data <- time_summary_roll %>%
  select(time_min,
         roll_mean_ch1, roll_mean_ch2, roll_mean_ch3,
         roll_sem_ch1, roll_sem_ch2, roll_sem_ch3) %>%
  pivot_longer(
    cols = -time_min,
    names_to = c(".value", "channel"),
    names_pattern = "roll_(mean|sem)_ch(\\d)"
  )

 plot_data$channel <- recode(
  plot_data$channel,
  norm_ch1 = "Channel 1",
  norm_ch2 = "Channel 2",
  norm_ch3 = "Channel 3"
)

 
ggplot(plot_data, aes(x = time_min, y = mean, color = channel, fill = channel)) +

  geom_ribbon(
    aes(ymin = mean - sem, ymax = mean + sem),
    alpha = 0.2,
    color = NA
  ) +

  geom_line(linewidth = 1.2) +

  scale_color_manual(
    values = c("1"="blue", "2"="green", "3"="red"),
    labels = c("Channel 1","Channel 2","Channel 3")
  ) +

  scale_fill_manual(
    values = c("1"="blue", "2"="green", "3"="red")
  ) +

  theme_bw() +

  labs(
    x = "Time (minutes)",
    y = "Δ Intensity (relative to start)",
    title = "Normalized channel dynamics"
  )

```

#Normalized time-size analysis
```{r}
summary_df <- time_summary_merge %>%
  group_by(Track.ID) %>%
  mutate(norm_frame = cell_Frame - min(cell_Frame, na.rm = TRUE)) %>%
  ungroup() %>%
  group_by(norm_frame) %>%
  summarise(
    mean_area = mean(cell_Area, na.rm = TRUE),
    sem_area = sd(cell_Area, na.rm = TRUE) / sqrt(sum(!is.na(cell_Area))),

    mean_nuc_area = mean(nuc_Area, na.rm = TRUE),
    sem_nuc_area = sd(nuc_Area, na.rm = TRUE) / sqrt(sum(!is.na(nuc_Area))),

    mean_nuc_circ = mean(nuc_Circ., na.rm = TRUE),
    sem_nuc_circ = sd(nuc_Circ., na.rm = TRUE) / sqrt(sum(!is.na(nuc_Circ.))),

    mean_mCherry = mean(cell_Mean.ch3, na.rm = TRUE),
    sem_mCherry = sd(cell_Mean.ch3, na.rm = TRUE) / sqrt(sum(!is.na(cell_Mean.ch3))),

    mean_GFP = mean(cell_Mean.ch2, na.rm = TRUE),
    sem_GFP = sd(cell_Mean.ch2, na.rm = TRUE) / sqrt(sum(!is.na(cell_Mean.ch2))),

    mean_BFP = mean(nuc_Mean.ch1, na.rm = TRUE),
    sem_BFP = sd(nuc_Mean.ch1, na.rm = TRUE) / sqrt(sum(!is.na(nuc_Mean.ch1))),

    n_area = sum(!is.na(cell_Area)),
    n_nuc_area = sum(!is.na(nuc_Area)),
    n_nuc_circ = sum(!is.na(nuc_Circ.)),
    n_mCherry = sum(!is.na(cell_Mean.ch3)),
    n_GFP = sum(!is.na(cell_Mean.ch2)),
    n_BFP = sum(!is.na(nuc_Mean.ch1)),
    .groups = "drop"
  )

ggplot(summary_df, aes(x = norm_frame, y = mean_area)) +
  geom_line() +
  geom_ribbon(aes(ymin = mean_area - sem_area,
                  ymax = mean_area + sem_area),
              alpha = 0.3) +
  labs(title = "normalized time - cell size", x = "normalized time", y = "Cell area")

ggplot(summary_df, aes(x = norm_frame, y = mean_nuc_area)) +
  geom_line() +
  geom_ribbon(aes(ymin = mean_nuc_area - sem_nuc_area,
                  ymax = mean_nuc_area + sem_nuc_area),
              alpha = 0.3) +
  labs(title = "normalized time - Nucleus size", x = "normalized time", y = "Nuc area")

ggplot(summary_df, aes(x = norm_frame, y = mean_nuc_circ)) +
  geom_line() +
  geom_ribbon(aes(ymin = mean_nuc_circ - sem_nuc_circ,
                  ymax = mean_nuc_circ + sem_nuc_circ),
              alpha = 0.3) +
  labs(title = "normalized time - Nucleus circularity", x = "normalized time", y = "Nuc circ")


ggplot(summary_df, aes(x = norm_frame, y = mean_mCherry)) +
  geom_line() +
  geom_ribbon(aes(ymin = mean_mCherry - sem_mCherry,
                  ymax = mean_mCherry + sem_mCherry),
              alpha = 0.3) +
  labs(title = "normalized time - mCherry signal", x = "normalized time", y = "mCherry")

ggplot(summary_df, aes(x = norm_frame, y = mean_GFP)) +
  geom_line() +
  geom_ribbon(aes(ymin = mean_GFP - sem_GFP,
                  ymax = mean_GFP + sem_GFP),
              alpha = 0.3) +
  labs(title = "normalized time - GFP signal", x = "normalized time", y = "GFP")

ggplot(summary_df, aes(x = norm_frame, y = mean_BFP)) +
  geom_line() +
  geom_ribbon(aes(ymin = mean_BFP - sem_BFP,
                  ymax = mean_BFP + sem_BFP),
              alpha = 0.3) +
  labs(title = "normalized time - BFP signal", x = "normalized time", y = "BFP")

# combined normalized plot: corrected GFP, BFP, mCherry on the same [0,1] axes
norm_pop <- time_summary_merge %>%
  group_by(`Track.ID`) %>%
  mutate(
    norm_frame = cell_Frame - min(cell_Frame, na.rm = TRUE),
    gfp_corr_norm = {
      r_min <- min(cell_Mean.ch2.corrected, na.rm = TRUE)
      r_max <- max(cell_Mean.ch2.corrected, na.rm = TRUE)
      if (r_max - r_min > 1e-9) (cell_Mean.ch2.corrected - r_min) / (r_max - r_min) else rep(NA_real_, n())
    },
    bfp_norm = {
      r_min <- min(nuc_Mean.ch1, na.rm = TRUE)
      r_max <- max(nuc_Mean.ch1, na.rm = TRUE)
      if (r_max - r_min > 1e-9) (nuc_Mean.ch1 - r_min) / (r_max - r_min) else rep(NA_real_, n())
    },
    red_norm = {
      r_min <- min(cell_Mean.ch3, na.rm = TRUE)
      r_max <- max(cell_Mean.ch3, na.rm = TRUE)
      if (r_max - r_min > 1e-9) (cell_Mean.ch3 - r_min) / (r_max - r_min) else rep(NA_real_, n())
    }
  ) %>%
  ungroup() %>%
  group_by(norm_frame) %>%
  summarise(
    mean_gfp_corr = mean(gfp_corr_norm, na.rm = TRUE),
    sem_gfp_corr  = sd(gfp_corr_norm,   na.rm = TRUE) / sqrt(sum(!is.na(gfp_corr_norm))),
    mean_bfp      = mean(bfp_norm,       na.rm = TRUE),
    sem_bfp       = sd(bfp_norm,         na.rm = TRUE) / sqrt(sum(!is.na(bfp_norm))),
    mean_red      = mean(red_norm,        na.rm = TRUE),
    sem_red       = sd(red_norm,          na.rm = TRUE) / sqrt(sum(!is.na(red_norm))),
    .groups = "drop"
  ) %>%
  arrange(norm_frame)

ggplot(norm_pop, aes(x = norm_frame)) +
  geom_ribbon(aes(ymin = mean_gfp_corr - sem_gfp_corr,
                  ymax = mean_gfp_corr + sem_gfp_corr),
              fill = "green3", alpha = 0.2) +
  geom_line(aes(y = mean_gfp_corr, color = "Corrected GFP"), linewidth = 1) +
  geom_ribbon(aes(ymin = mean_bfp - sem_bfp,
                  ymax = mean_bfp + sem_bfp),
              fill = "dodgerblue", alpha = 0.2) +
  geom_line(aes(y = mean_bfp, color = "BFP (nuclear)"), linewidth = 1) +
  geom_ribbon(aes(ymin = mean_red - sem_red,
                  ymax = mean_red + sem_red),
              fill = "tomato", alpha = 0.2) +
  geom_line(aes(y = mean_red, color = "mCherry"), linewidth = 1) +
  scale_color_manual(values = c(
    "Corrected GFP" = "green3",
    "BFP (nuclear)" = "dodgerblue",
    "mCherry"       = "tomato"
  )) +
  labs(
    title    = "Population mean normalized channel dynamics",
    subtitle = "Each cell normalized 0-1 per channel to its own min/max; mean +/- SEM",
    x     = "Normalized frame (from track start)",
    y     = "Normalized intensity (0-1)",
    color = NULL
  ) +
  theme_bw() +
  theme(legend.position = "top")

```

#Correlations
```{r}
#Crosscorrelations

#Try chunks

plot_avg_ccf(data = time_summary_merge, feature_x = "nuc_Mean.ch1", feature_y = "cell_Mean.ch2", track_col = "Track.ID", time_col = "cell_Frame", lag_max = 20, min_points = 10, zscore = TRUE)
plot_avg_ccf(data = time_summary_merge, feature_x = "cell_Area", feature_y = "nuc_Area", track_col = "Track.ID", time_col = "cell_Frame", lag_max = 20, min_points = 10, zscore = TRUE)
plot_avg_ccf(data = time_summary_merge, feature_x = "nuc_Circ.", feature_y = "nuc_Area", track_col = "Track.ID", time_col = "cell_Frame", lag_max = 20, min_points = 10, zscore = TRUE)
plot_avg_ccf(data = time_summary_merge, feature_x = "cell_speed", feature_y = "cell_Area", track_col = "Track.ID", time_col = "cell_Frame", lag_max = 20, min_points = 10, zscore = TRUE)
#ccf_res$plot #ccf_res$summary #ccf_res$per_track

#Try crossvalidation (leave 10% out each time)
#elasticnet - linear model
#tabiclv2 instead of XGboost
#zscore normalization for every feature
#coefficient of variation (delta normalized to mean/to start), variance within the time frame, zscore normalized delta
#deep learning using the signals themselves and give time to red as labels.

#Static feature correlations 
draw_it(time_summary_plot, "time_to_green_start", "delay_green_to_red_min")
draw_it(time_summary_plot, "time_to_green_start", "delay_green_to_red_min")
draw_it(time_summary_plot, "delay_green_to_blue_min", "delay_green_to_red_min")


```
#Features table
```{r}

early_df <- spots3 %>%
  arrange(`Track.ID`, Frame) %>%
  group_by(`Track.ID`) %>%
  mutate(norm_frame = Frame - min(Frame, na.rm = TRUE)) %>%
  filter(norm_frame < 16) %>%
  ungroup() %>%
  mutate(
    nuc_cell_area_ratio = ifelse(
      is.na(cell_Area) | cell_Area == 0,
      NA_real_,
      nuc_Area / cell_Area
    )
  )

early_df <- early_df %>%
  mutate(
    across(
      where(~ is.character(.) | is.factor(.)),
      ~ suppressWarnings(as.numeric(as.character(.)))
    )
  )

feature_table <- early_df %>%
  group_by(`Track.ID`) %>%
  summarise(
    # --------------------------------
    # cell size
    # --------------------------------
    cell_area_start = safe_first(cell_Area),
    cell_area_mean  = mean(cell_Area, na.rm = TRUE),
    cell_area_sd    = safe_sd(cell_Area),
    cell_area_slope = safe_slope(cell_Area, norm_frame),

    # --------------------------------
    # nucleus size / shape
    # --------------------------------
    nuc_area_start = safe_first(nuc_Area),
    nuc_area_mean  = mean(nuc_Area, na.rm = TRUE),
    nuc_area_slope = safe_slope(nuc_Area, norm_frame),

    nuc_circ_mean  = mean(nuc_Circ., na.rm = TRUE),
    nuc_circ_sd    = safe_sd(nuc_Circ.),

    # --------------------------------
    # nucleus / cell size ratio
    # --------------------------------
    nuc_cell_area_ratio_start = safe_first(nuc_cell_area_ratio),
    nuc_cell_area_ratio_mean  = mean(nuc_cell_area_ratio, na.rm = TRUE),
    nuc_cell_area_ratio_slope = safe_slope(nuc_cell_area_ratio, norm_frame),

    # --------------------------------
    # motion
    # --------------------------------
    speed_mean  = mean(cell_speed, na.rm = TRUE),
    speed_sd    = safe_sd(cell_speed),
    speed_slope = safe_slope(cell_speed, norm_frame),

    # --------------------------------
    # GFP / reporter features (channel 2, bleedthrough-corrected)
    # --------------------------------
    gfp_start = safe_first(cell_Mean.ch2.corrected),
    gfp_mean  = mean(cell_Mean.ch2.corrected, na.rm = TRUE),
    gfp_sd    = safe_sd(cell_Mean.ch2.corrected),
    gfp_slope = safe_slope(cell_Mean.ch2.corrected, norm_frame),

    gfp_snr_mean = mean(cell_SNR.ch2, na.rm = TRUE),
    gfp_snr_sd   = safe_sd(cell_SNR.ch2),

    # --------------------------------
    # brightfield features (channel 4)
    # --------------------------------
    bf_snr_mean   = mean(cell_SNR.ch4, na.rm = TRUE),
    bf_ctrst_mean = mean(cell_Ctrst.ch4, na.rm = TRUE),
    bf_ctrst_sd   = safe_sd(cell_Ctrst.ch4),

    # --------------------------------
    # morphology descriptors
    # --------------------------------
    solidity_mean    = mean(cell_Solidity, na.rm = TRUE),
    solidity_sd      = safe_sd(cell_Solidity),

    shape_index_mean = mean(cell_Shape.index, na.rm = TRUE),

    long_axis_mean   = mean(cell_El..long.axis, na.rm = TRUE),
    long_axis_sd     = safe_sd(cell_El..long.axis),
    long_axis_slope  = safe_slope(cell_El..long.axis, norm_frame),

    .groups = "drop"
  ) %>%
  mutate(
    across(
      where(is.numeric),
      ~ replace(., is.infinite(.), NA)
    )
  )

feature_table_raw <- feature_table

feature_table_z <- feature_table %>%
  mutate(
    across(
      .cols = where(is.numeric) & !matches("^Track\\.ID$"),
      .fns = zscore_feature,
      .names = "{.col}_z"
    )
  ) %>%
  select(`Track.ID`, ends_with("_z"))


cor_mat <- cor(
  feature_table_z[, -which(names(feature_table_z) == "Track.ID")],
  use = "pairwise.complete.obs",
  method = "pearson"
)

corrplot(cor_mat, method = "color", tl.cex = 0.6)

feat_tmp = merge(feature_table_z, time_summary, by = "Track.ID")

cors <- sapply(feat_tmp[, 2:31], function(col) {
  x <- suppressWarnings(as.numeric(as.character(col)))
  y <- suppressWarnings(as.numeric(as.character(feat_tmp$delay_green_to_red)))
  
  keep <- is.finite(x) & is.finite(y)
  
  if (sum(keep) < 3) return(NA_real_)
  if (sd(x[keep]) == 0 || sd(y[keep]) == 0) return(NA_real_)
  
  cor(x[keep], y[keep])
})

cor_df <- data.frame(
  feature = names(cors),
  correlation = as.numeric(cors)
) %>%
  filter(!is.na(correlation))

ggplot(data=cor_df, aes(x=feature, y=correlation, fill = feature)) + geom_bar(stat="identity") + theme(axis.title.x=element_blank(), axis.text.x=element_blank(), axis.ticks.x=element_blank())


draw_it(feat_tmp, "gfp_mean_z", "delay_green_to_red_min")

cor(feat_tmp[, c("gfp_mean_z", "gfp_slope_z", "gfp_sd_z")], use = "pairwise.complete.obs")

ggplot(feat_tmp, aes(x = delay_green_to_red_min)) + geom_density(fill = "red") + labs(title = "green to red delay (minutes)")

df_surv <- feat_tmp %>%
  mutate(
    event = ifelse(is.finite(delay_green_to_red_min), 1, 0),
    time = ifelse(is.finite(delay_green_to_red_min),
                  delay_green_to_red_min,
                  max(delay_green_to_red_min))
  )


fit <- survival::coxph(
  survival::Surv(time, event) ~ gfp_mean_z + cell_area_sd_z,
  data = df_surv
)
summary(fit)

```
Linear model
```{r}

df_lm = feat_tmp[,c(2:31,49)]

df_lm[is.na(df_lm$delay_green_to_red_min),"delay_green_to_red_min"] = 4500

df_lm_imp <- df_lm %>%
  mutate(
    across(
      where(is.numeric) & !matches("delay_green_to_red_min"),
      ~ ifelse(is.na(.), median(., na.rm = TRUE), .)
    )
  )

lm_model <- lm(delay_green_to_red_min ~ ., data = df_lm_imp)
summary(lm_model)

#regularized regression

# LASSO
x <- model.matrix(delay_green_to_red_min ~ ., df_lm_imp)[, -1]
y <- na.omit(df_lm_imp$delay_green_to_red_min)

cv_model <- cv.glmnet(x, y, alpha = 1)

best_lambda <- cv_model$lambda.min

lasso_model <- glmnet(x, y, alpha = 1, lambda = best_lambda)

coef(lasso_model)

# ElasticNet
cv_model_en <- cv.glmnet(x, y, alpha = 0.5)

best_lambda_en <- cv_model_en$lambda.min

enet_model <- glmnet(x, y, alpha = 0.5, lambda = best_lambda_en)

coef(enet_model)
plot(cv_model_en)

selected_features <- replicate(50, {
  idx <- sample(1:nrow(x), replace = TRUE)
  cv <- cv.glmnet(x[idx,], y[idx], alpha = 1)
  model <- glmnet(x[idx,], y[idx], alpha = 1, lambda = cv$lambda.min)
  rownames(coef(model))[which(coef(model) != 0)]
})

bootstrap_feats = as.data.frame(table(unlist(selected_features)))


bootstrap_feats = bootstrap_feats[-1,]
bootstrap_feats <- bootstrap_feats %>%
  arrange(desc(Freq))

ggplot(bootstrap_feats, aes(x = reorder(Var1, Freq), y = Freq, fill = Var1)) +
  geom_bar(stat = "identity") +
  coord_flip() + theme(legend.position = "none") + labs(title = "informative features in bootstrap (n = 50)")


#Turn red or not
df_cls <- feat_tmp[,c(2:31,49)] %>%
  mutate(
    turns_red = ifelse(is.finite(delay_green_to_red_min), 1, 0)
  )

df_cls <- df_cls %>%
  mutate(
    across(where(is.numeric), ~ ifelse(is.infinite(.), NA, .))
  ) %>%
  select(-delay_green_to_red_min)

df_cls_imp <- df_cls %>%
  mutate(
    across(
      where(is.numeric) & !matches("turns_red"),
      ~ ifelse(is.na(.), median(., na.rm = TRUE), .)
    )
  )

lm_model <- lm(turns_red ~ ., data = df_cls_imp)
summary(lm_model)

glm_model <- glm(turns_red ~ ., data = df_cls_imp, family = binomial)
summary(glm_model)


x <- model.matrix(turns_red ~ ., df_cls_imp)[, -1]
y <- df_cls_imp$turns_red

cv <- cv.glmnet(x, y, alpha = 1, family = "binomial")

lasso_model <- glmnet(x, y, alpha = 1, lambda = cv$lambda.min, family = "binomial")

coef(lasso_model)

```

#Cox ElasticNet + KM risk stratification
```{r}

# ── proper survival outcome ────────────────────────────────────────────────────
# For censored cells use actual track duration from green onset, not imputed max.
track_duration_min <- spots3 %>%
  group_by(`Track.ID`) %>%
  summarise(
    track_dur_min = (max(`T..sec.`, na.rm = TRUE) - min(`T..sec.`, na.rm = TRUE)) / 60,
    .groups = "drop"
  )

df_cox <- feat_tmp %>%
  left_join(track_duration_min, by = "Track.ID") %>%
  mutate(
    event     = as.integer(is.finite(delay_green_to_red_min)),
    surv_time = ifelse(event == 1, delay_green_to_red_min, track_dur_min)
  ) %>%
  filter(is.finite(surv_time), surv_time > 0)

cat(sprintf("Cox: %d cells | %d events | %d censored\n",
            nrow(df_cox), sum(df_cox$event), sum(df_cox$event == 0)))

# ── feature matrix (z-scored, same columns as linear model) ───────────────────
feat_cols_cox <- names(df_cox)[grepl("_z$", names(df_cox))]

X_cox <- df_cox %>%
  select(all_of(feat_cols_cox)) %>%
  mutate(across(everything(), ~ ifelse(is.na(.), median(., na.rm = TRUE), .))) %>%
  as.matrix()

surv_y <- survival::Surv(df_cox$surv_time, df_cox$event)

# ── Cox ElasticNet (alpha = 0.5) ───────────────────────────────────────────────
set.seed(42)
cv_cox   <- cv.glmnet(X_cox, surv_y, family = "cox", alpha = 0.5, nfolds = 10)
cox_en   <- glmnet(X_cox, surv_y, family = "cox", alpha = 0.5, lambda = cv_cox$lambda.min)

plot(cv_cox, main = "Cox ElasticNet CV (alpha=0.5, 10-fold)")

# coefficients
cox_coefs <- as.data.frame(as.matrix(coef(cox_en))) %>%
  rename(log_HR = s0) %>%
  filter(log_HR != 0) %>%
  arrange(desc(abs(log_HR)))

print(cox_coefs)

# concordance
risk_score <- as.numeric(predict(cox_en, newx = X_cox, s = cv_cox$lambda.min, type = "link"))
conc       <- survival::concordance(surv_y ~ risk_score)
# glmnet Cox: higher score = higher hazard; concordance < 0.5 means sign is flipped
c_stat <- max(conc$concordance, 1 - conc$concordance)
cat(sprintf("C-statistic: %.3f\n", c_stat))

# ── bootstrap feature stability ────────────────────────────────────────────────
set.seed(99)
boot_cox <- replicate(50, {
  idx <- sample(nrow(X_cox), replace = TRUE)
  cv  <- cv.glmnet(X_cox[idx, ], surv_y[idx], family = "cox", alpha = 0.5, nfolds = 5)
  m   <- glmnet(X_cox[idx, ], surv_y[idx], family = "cox", alpha = 0.5, lambda = cv$lambda.min)
  cc  <- as.matrix(coef(m))
  rownames(cc)[which(cc != 0)]
}, simplify = FALSE)

cox_boot_df <- as.data.frame(table(unlist(boot_cox))) %>%
  rename(feature = Var1, freq = Freq) %>%
  arrange(desc(freq)) %>%
  head(20)

ggplot(cox_boot_df, aes(x = reorder(feature, freq), y = freq, fill = feature)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  theme_bw() + theme(legend.position = "none") +
  labs(title = sprintf("Cox bootstrap feature stability (n=50)  |  C=%.3f", c_stat),
       x = NULL, y = "Times selected out of 50") +
  geom_hline(yintercept = 25, linetype = "dashed", color = "red")

# ── coefficient plot ───────────────────────────────────────────────────────────
ggplot(cox_coefs %>% tibble::rownames_to_column("feature"),
       aes(x = reorder(feature, log_HR), y = log_HR,
           fill = ifelse(log_HR > 0, "faster to red", "slower to red"))) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("faster to red" = "tomato", "slower to red" = "steelblue")) +
  theme_bw() +
  labs(title = sprintf("Cox ElasticNet coefficients  (C=%.3f)", c_stat),
       x = NULL, y = "log Hazard Ratio (per SD)",
       fill = NULL)

# ── KM curves: cross-validated risk groups (out-of-sample) ────────────────────
# Train Cox ElasticNet on k-1 folds, predict on held-out fold → repeat.
# Risk groups are assigned only from held-out predictions, so KM separation
# is truly out-of-sample and not inflated by in-sample fitting.

k_folds  <- 5
set.seed(42)
fold_ids <- sample(rep(1:k_folds, length.out = nrow(df_cox)))

cv_risk_scores <- numeric(nrow(df_cox))

for (fold in seq_len(k_folds)) {
  test_idx  <- which(fold_ids == fold)
  train_idx <- which(fold_ids != fold)

  X_train <- X_cox[train_idx, , drop = FALSE]
  X_test  <- X_cox[test_idx,  , drop = FALSE]
  y_train <- surv_y[train_idx]

  # select lambda by inner CV on training fold
  inner_cv <- cv.glmnet(X_train, y_train, family = "cox",
                         alpha = 0.5, nfolds = 5)
  fold_model <- glmnet(X_train, y_train, family = "cox",
                        alpha = 0.5, lambda = inner_cv$lambda.min)

  cv_risk_scores[test_idx] <- as.numeric(
    predict(fold_model, newx = X_test, type = "link")
  )
}

# concordance on out-of-sample scores
conc_cv   <- survival::concordance(surv_y ~ cv_risk_scores)
c_stat_cv <- max(conc_cv$concordance, 1 - conc_cv$concordance)
cat(sprintf("Cross-validated C-statistic: %.3f\n", c_stat_cv))

# median split on out-of-sample scores
df_cox <- df_cox %>%
  mutate(
    risk_score_cv = cv_risk_scores,
    risk_group_cv = ifelse(cv_risk_scores > median(cv_risk_scores),
                           "High risk", "Low risk")
  )

km_fit_cv <- survival::survfit(surv_y ~ df_cox$risk_group_cv)

plot(km_fit_cv,
     col = c("tomato", "steelblue"), lwd = 2,
     xlab = "Time from green onset (min)", ylab = "P(not yet red)",
     main = sprintf("KM curves — cross-validated risk groups (%d-fold)  |  C=%.3f",
                    k_folds, c_stat_cv))
legend("topright", bty = "n",
       legend = c(sprintf("High risk  (n=%d)", sum(df_cox$risk_group_cv == "High risk")),
                  sprintf("Low risk   (n=%d)", sum(df_cox$risk_group_cv == "Low risk"))),
       col = c("tomato", "steelblue"), lwd = 2)

# log-rank test p-value
logrank <- survival::survdiff(surv_y ~ df_cox$risk_group_cv)
p_val   <- 1 - pchisq(logrank$chisq, df = 1)
cat(sprintf("Log-rank p-value: %.4f\n", p_val))

```

#PCA
```{r}

feature_mat <- feature_table_z %>%
  select(where(is.numeric))


#impute
feature_mat <- feature_mat %>%
  mutate(
    across(
      where(is.numeric),
      ~ ifelse(is.na(.), median(., na.rm = TRUE), .)
    )
  )

feature_mat <- feature_mat %>%
  mutate(across(everything(), ~ ifelse(is.finite(.), ., NA_real_)))

# drop all-NA columns
keep_not_all_na <- sapply(feature_mat, function(x) !all(is.na(x)))
feature_mat <- feature_mat[, keep_not_all_na, drop = FALSE]

feature_mat_complete <- feature_mat %>%
  filter(if_all(everything(), ~ is.finite(.)))

feature_mat_complete = feature_mat_complete[, -which(names(feature_mat_complete) == "Track.ID")]

pca_res <- prcomp(
  feature_mat_complete,
  center = TRUE,
  scale. = TRUE
)

var_explained <- pca_res$sdev^2 / sum(pca_res$sdev^2)

plot(var_explained, type = "b",
     xlab = "PC",
     ylab = "Variance explained")

pc_scores <- as.data.frame(pca_res$x)

pc_scores$Track.ID <- feature_table$Track.ID[
  as.numeric(rownames(pc_scores))
]

pc_scores <- pc_scores %>%
  left_join(time_summary %>%
              mutate(turned_red = is.finite(time_to_red_start)),
            by = "Track.ID")


ggplot(pc_scores, aes(x = PC1, y = PC2, color = pc_scores$time_to_green_start_min)) +
  geom_point(alpha = 0.7) + labs(x = paste0("PC1 (", round(var_explained[1], 2), ")"), y = paste0("PC2 (", round(var_explained[2], 2), ")")) + theme_bw()

loadings <- as.data.frame(pca_res$rotation)

loadings$PC1 %>% sort(decreasing = TRUE) %>% head(10)
loadings$PC1 %>% sort() %>% head(10)

X <- pc_scores %>%
  select(starts_with("PC")) %>%
  as.matrix()

X <- pc_scores %>%
  select(PC1:PC10) %>%
  as.matrix()
```








